---
author: "Daniel Fabio Groth"
title: "Assignment 4"
format: html
editor: visual
warning: false
echo: false
message: false
toc: true
toc-title: "Innholdsliste"
---

```{r}
rm (list = ls())
library(tidyverse)
library(readr)
```

# Task 1

## Data

You will find the data for these tasks at the following url:

[https://raw.githubusercontent.com/uit-sok-1005-v23/uit-sok-1005-v23.github.io/main/storedata.csv.](https://raw.githubusercontent.com/uit-sok-1005-v23/uit-sok-1005-v23.github.io/main/storedata.csv)

The data file contains 7185 observations and 9 variables.

1.  For the last 3 months of 2017, calculate the total Sales by month, for Region 1 and Region 9 in the Customer_Segment, Corporate, and Consumer. This output is Table 1.

```{r}
storedata <- read_csv("https://raw.githubusercontent.com/uit-sok-1005-v23/uit-sok-1005-v23.github.io/main/storedata.csv", show_col_types = FALSE)
```

```{r}
storedata %>% 
  filter(Region %in% c("Region 1", "Region 9"), Customer_Segment %in% c("Corporate", "Consumer")) %>% 
  group_by(Month, Region, Customer_Segment) %>% 
  summarise(Total_Sales = sum(Sales)) %>% 
  knitr::kable()

```

2: Make a plot of the monthly total Sales in Region 1 and Region 13 in 2015, 2016, and 2017. This output is Figure 1.

```{r}

```

3: In Figure 1, identify the months where the total Sales in Region 13 is greater than the total Sales in Region 1. This output is Table 2.

```{r}

```

4: Find the average Profit per Customer_Segment and Product_Category in 2017, for all regions except Region 3, 5 and 8. What segment produced the highest average profit? This output is Table 3.

```{r}

```

# Task 2.

In this task, feel free to use any API or package/library that downloads the data to your session. Use code and download daily stock prices for Exxon Mobil Corporation (XOM), traded at NYSE. The Yahoo! Finance site is a convenient place to find [the data](https://finance.yahoo.com/quote/XOM?p=XOM&.tsrc=fin-srch&guccounter=1&guce_referrer=aHR0cHM6Ly93d3cuZ29vZ2xlLmNvbS8&guce_referrer_sig=AQAAAFPxsq3OMyVmjUi0J0prTIqCvNsFgCcjQ5H0WY8yNWD_IC8QU8uW4L99CYtwUeppCqHZMSBzoYKPc0DEbX-JwH7pTqwlQuTTxACSjuGme82tlaS44Ssq2RtiMmi4wd-AMN63-1UJrlWVNm_gaRpKyCHWmzH7O0jufICxvKb1qyYx). Use the Adjusted closing price from January 4th 2010 as the starting date. And calculate the monthly average using trading volume as a weight, and save this variable as "exxon".

```{r}

```

Use code to download the daily Brent Crude Oil Price from [FRED](https://fred.stlouisfed.org/series/DCOILBRENTEU) from January 4th 2010 as the starting date. And calculate the monthly arithmetic average. Save the monthly arithmetic average values as "oil". In both variables, take December 2022, or 2022:12 for shorthand as a last data point. Plot both variables, i.e., exxon and oil " in the same plot window. Here, the x-axis should be the "date" variable. Comment on the plots. Now take "exxon" as a y-variable and "oil" as an x-variable.

```{r}


```

Use R's [lm() function](https://www.rdocumentation.org/packages/stats/versions/3.6.2/topics/lm). Set the variable on the y-axis and x-axis, and specify the data set.

*LM(*<Y variable name> *\~* <X variable>*, data=*<dataset name>*)*

After "running" the code, how do you interpret the estimated coefficients? 

Skriv svar her
